---
title: 'Supplementary material of "Estimating quantile treatment effect on the original scale of the outcome variable: a case study of common cold treatments"' 
author: | 
 | Harri Hemilä and Matti Pirinen 
 | University of Helsinki
output:
  pdf_document: default
  html_document: default
date: "2023-06-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("bqte_functions.R")
```


# 1. Interpretation of BQTE

Back-transformed quantile treatment effect (BQTE) is a function that describes
the difference between the distributions of outcome values of two populations.
We will call the two populations that we compare 
as the **treatment** population and the **control** population, 
motivated by the need to compare how
a particular treatment affects the outcome value in a randomized controlled trial (RCT).
In an RCT, the study sample is randomly split into the treatment group (that represents
the treatment population) and the control group (that represents the control population).
The treatment of interest is applied to the treatment group and placebo is applied
to the control group. The possible difference in the outcome distributions between
the treatment group and the control group is interpreted as the population-level treatment effect.
We will use the BQTE function to quantify this 
difference between the two populations.

For any possible outcome value $x$, the theoretical value of $\text{BQTE}(x)$ is defined as follows:

 1. Find which proportion of the control population has outcome values $\leq x$. Call this
 proportion $F(x)$.
 
 2. Find what is the corresponding outcome value $x^*$ in the treatment population for which
 the proportion of the treatment population having outcome values $\leq x^*$ is the same $F(x)$.

 3. Define $\text{BQTE}(x) = x^* - x.$
 
Thus, $\text{BQTE}$ describes the difference between the outcome value
of the treatment population and the control population when this comparison
is done between treated and control individuals whose outcome values
rank similarly among their respective populations.
As the value of BQTE may vary across outcome values, we present $\text{BQTE}$ as 
a function of outcome value $x$ in the control group.

Mathematically,
$$\text{BQTE}(x) = {F^*}^{-1}\left(F(x)\right) -x,$$
where $F$ and $F^{*}$ are the cumulative distribution functions of the outcome values of the
control and treatment populations, respectively, and ${F^*}^{-1}$ is the inverse function of ${F^*}$. 

The difference in the means of the populations, called the **average treatment effect** (ATE), 
is the most common approach to assess the effect of the treatment at the population level.
However, the BQTE function measures the difference between the two populations in 
a more detailed way than the single value of ATE. 
For example, if we observe that BQTE is not constant across the outcome values, then we can conclude that
the individual-level treatment effect (see below for definition) cannot be constant across individuals.
Or, if the relative BQTE (defined as $\text{BQTE}(x)/x$) is not constant, then we know that
the individual-level treatment effect cannot be proportional to the untreated 
outcome value by a single constant.
Neither of these pieces of information can be inferred from the value of the ATE alone.
Next we consider what additional information we can extract from the BQTE function.


### 1.1. When does BQTE estimate an individual-level treatment effect?

We define two instances of the concept of **treatment effect**.
A **population-level treatment effect** is the difference
between the outcome distributions of the treatment and control populations and it can be
estimated from RCT data by different approaches such as by ATE or BQTE which contain different
levels of information about the population-level treatment effect.
Instead, an **individual-level treatment effect** is a theoretical construction
that cannot, in general, be estimated from an RCT without some additional assumptions. 
To formally define the individual-level treatment effect,
we follow the framework of the potential outcomes by Rubin 
(1974; Journal of Educational Psychology, 66(5), 688–701).
Suppose that we have not yet carried out the experiment. 
For any individual $i$, we can imagine two potential outcome variables:

 * $X_i$ would be the outcome variable if individual $i$ belonged to the control population,
 * $X_i^{*}$ would be the outcome variable if individual $i$ belonged to the treatment population.

Before the experiment, we consider both $X_i$ and $X_i^{*}$ as random variables and 
their difference, $T_i = X_i^{*} - X_i$, defines a random variable that 
represents the theoretical treatment effect for individual $i$. 
Once we do the experiment, we will observe a value for either
$X_i$ (if $i$ was assigned to the control group) or $X_i^{*}$ 
(if $i$ was assigned to the treatment group) but never for both $X_i$ and $X_i^{*}$ simultaneously.
Thus, the experiment does not provide a direct way to estimate the treatment 
effect $T_i$ of individual $i$. 
Note that from now on, we will refer by the random variables $X$ and $X^*$
to the untreated and treated potential outcomes of the *same* individual
even when we do not specify that same individual by using a common subscript as above.

To conceptually connect BQTE, that quantifies the population-level treatment effect, 
to the individual-level treatment effects, 
we will consider the concept of order-preserving treatments.

 **Definition.** We say that a **treatment preserves order** (or that the **treatment
 is order-preserving**) if the untreated outcome values 
 of a set of individuals have, on average, the same relative order as the
 treated outcome values of the same set of individuals would have had, 
 if these individuals had been treated. 
 Formally, we define that the treatment preserves order when, for every outcome value $x$,
 $F^*\left(\text{E}(X^*\,|\,X=x)\right) = F(x)$.
 
For example, consider a trial where the outcome is the duration of the common cold, 
the treatment group is treated with zinc acetate lozenges and the 
control group is given placebo. 
The assumption that the treatment preserves order
means that the durations of the colds of the control individuals 
would have had approximately the same relative order if those individuals had been 
treated with zinc acetate lozenges rather than given placebo. 

While, based on RCT data, we cannot directly confirm that a treatment preserves order,
we can make the following remarks. 

 * First, the assumption that the treatment preserves order seems more plausible the more
homogeneous the individuals in the study population are, because within a homogeneous
population there are less differences between the individuals that could affect 
how the treatment applies to each individual. 
Thus, for example, the assumption of order-preserving treatment seems more plausible 
in an RCT that studies females between 40-45 years of age
than in another RCT that studies both males and females between 18-80 years of age.

 * Second, in case that some covariate information were available on the individuals included
in the RCT (such as sex, age, medication, disease history etc.),
then we would expect that under an order-preserving treatment
the way how the covariates predict the outcome values would be 
similar in the treatment population as it is in the control population.
For example, if male individuals tend to have longer disease duration than female individuals
in the control group,
then the same should be true in the treatment group if the treatment preserves order.
If, on the other hand, we observe that the covariates such as sex do not predict
the disease duration consistently in both the control and the treatment populations, 
then the assumption of order-preserving treatment may not be appropriate. 
Note that the treatment could preserve order in a subpopulation (such as in males only)
even if it did not preserve order in a larger population (such as in both sexes together).

The connection between the order-preserving treatments and the individual-level treatment effects
can be formulated as follows:
**BQTE can be interpreted as
the expected value of the individual-level treatment effect if and only if
the treatment preserves order.**

To justify the claim, consider a treatment that preserves order, that is, 
$F^*\left(\text{E}(X^*\,|\,X=x)\right) = F(x)$ for all $x$.
By applying ${F^*}^{-1}$ on both sides of the equation we get 
$$\text{E}(X^*\,|\,X=x) = {F^*}^{-1}\left(F(x)\right)$$
and since by definition $\text{BQTE}(x) = {F^*}^{-1}(F(x)) -x$ it follows that 
$$\text{BQTE}(x) = \text{E}(X^*\,|\, X = x) - x = \text{E}(X^* -X\,|\, X=x).$$
Thus, BQTE is the conditional expectation of the individual-level treatment effect and
we can interpret BQTE as an estimator of the individual-level treatment effect.

On the other hand, suppose that the treatment does not preserve order.
Then there exists some outcome value $x$ in the control population 
whose quantile level $F(x)$ does not remain constant under the treatment:
${F^*}\left(\text{E}(X^*\,|\, X = x)\right) \neq F(x).$
By applying ${F^*}^{-1}$ on both sides of this inequality, and following the steps
above, we see that
$$\text{BQTE}(x) \neq \text{E}(X^*\,|\, X = x) - x = \text{E}(X^* - X\,|\, X = x).$$
Thus, if the treatment does not preserve order, 
$\text{BQTE}(x)$ does not match with the conditional expectation of the 
individual-level treatment effect.

Theoretically, BQTE can be interpreted as the expected value of the individual-level 
treatment effect if and only if the treatment preserves order.
Unfortunately, there is no general way to verify whether a 
given treatment preserves order based on RCT data.
However, we can extract useful information about the individual-level 
treatment effects from the BQTE function even 
without an assumption about the preservation of order.

### 1.2 Bounds on the individual-level treatment effect 

In this section, we compute upper and lower bounds on the individual-level treatment effect
and further extend these bounds to the average treatment effect in the tails of the distribution.

Consider a treatment that does not necessarily preserve order. 
Let us denote the range of the possible value of the treatment population by 
$(L^*,U^*)$, where $L^*$ and $U^*$ refer, respectively, 
to the lower and upper bounds of the treatment population. 
We use the following estimators for these bounds: 
\begin{eqnarray*}
\widehat{L^*} &=& \min\{x^* \text{ of the treatment group}\}, \\
\widehat{U^*} &=& \max\{x^* \text{ of the treatment group}\},
\end{eqnarray*}
that is, we use the observed minimum and maximum of the outcome values
of the treatment group as the estimates of $L^*$ and $U^*$, respectively.
This leads to an estimated lower bound of $\widehat{L^*} - x$ and an estimated 
upper bound of $\widehat{U^*} - x$ for the treatment effect
of an untreated individual with an outcome value $X = x$.
The logic is that since we expect that (almost) all outcome values in the treatment
population are within the interval $\left[\widehat{L^*}, \widehat{U^*}\right]$, 
we expect that the treatment effect 
at $X = x$ is (almost always) bounded within $\left[\widehat{L^*} - x, \widehat{U^*} - x\right]$.
The accuracy of these bounds increases with the sample size of the treatment group
and decreases with increasing kurtosis (fatness of the tails) of the outcome distribution of
the treatment population.

We can extend the idea of the bounds to the average treatment effect of 
the upper or lower tail as follows.
For the outcome value $x$, 
define the **lower tail average treatment effect (LTATE)** as
$$\text{LTATE}(x) = \text{E}\left(X^* \,|\, X \leq x \right) - \text{E}\left(X \,|\, X \leq x \right).$$
and the **upper tail average treatment effect (UTATE)** as
$$\text{UTATE}(x) = \text{E}\left(X^* \,|\, X \geq x \right) - \text{E}\left(X \,|\, X \geq x \right).$$
While we cannot directly estimate these quantities for a general treatment, we can get
a lower bound for LTATE by
**lower tail back-transformed quantile treatment effect (LTBQTE)** and 
an upper bound for UTATE by 
**upper tail back-transformed quantile treatment effect (UTBQTE)**:
\begin{eqnarray*}
\text{LTBQTE}(x) &=& \text{E}\left(X^* \,|\, X^* \leq {F^*}^{-1}(F(x)) \right) - \text{E}\left(X \,|\, X \leq x \right)\\
\text{UTBQTE}(x) &=& \text{E}\left(X^* \,|\, X^* \geq {F^*}^{-1}(F(x)) \right) - \text{E}\left(X \,|\, X \geq x \right).
\end{eqnarray*}

Thus, LTBQTE (UTBQTE) is the difference in the averages of the treatment and control populations when
the average is computed for the same proportion of the populations from the lower (upper) tail.
Note that LTBQTE($x$) approaches ATE as $x$ gets large while UTBQTE($x$) approaches ATE as $x$ gets small.
Simple estimators to estimate $\text{LTBQTE}(x)$ and 
$\text{UTBQTE}(x)$ result by replacing the expectations by the 
numerical averages.

For these quantities, the following relations apply: 
$$\text{LTATE}(x) \geq \text{LTBQTE}(x) \quad \text{ and } \quad \text{UTATE}(x) \leq \text{UTBQTE}(x).$$

To verify the argument, let us consider the first inequality. 
We will show that $\text{LTATE}(x)  \geq \text{LTBQTE}(x),$ which equals to showing that
$$\text{E}\left(X^* \,|\, X \leq x \right) \geq \text{E}\left(X^* \,|\, X^* \leq {F^*}^{-1}(F(x)) \right).$$

Both quantities in the above inequality are integrals in the marginal space of random variable $X^*$.
The left-hand side is over the event $A =\{X \leq x\}$ and the right-hand side is over the event
$A^* = \{X^* \leq {F^*}^{-1}(F(x))\}$. 
Let $\mu$ be the underlying probability measure
of the two-dimensional random variable $(X,X^*)$. 
By definition of the marginal CDFs $F$ of $X$ and $F^*$ of $X^*$, both events have the same 
probability $\mu(A) = \mu(A^*) = F(x).$ 
Intuitively, among all events with the same 
probability, the event $A^*$ is the one with the smallest expected value of $X^*$.
To formally prove this, note that for all $\omega \in A \backslash A^*$, 
$X^*(\omega) > {F^*}^{-1}(F(x))$ and hence
\begin{eqnarray*}
\text{E}\left(X^* \,|\, X \leq x \right) &=& \int_{A}x^*d\mu(x^*) \\
&=& \int_{A\cap A^*}x^*d\mu(x^*) +  \int_{A\backslash A^*}x^*d\mu(x^*) \\
&\geq& \int_{A\cap A^*}x^*d\mu(x^*) +  \mu(A \backslash A^*) \cdot {F^*}^{-1}(F(x)) \\
&=& \int_{A\cap A^*}x^*d\mu(x^*) +  \mu(A^* \backslash A) \cdot {F^*}^{-1}(F(x)) \\
&\geq& \int_{A\cap A^*}x^*d\mu(x^*) + \int_{A^*\backslash A}x^*d\mu(x^*) \\
&=& \int_{A^*}x^*d\mu(x^*)\\
&=&\text{E}\left(X^* \,|\, X^* \leq {F^*}^{-1}(F(x)) \right).
\end{eqnarray*}

As a conclusion, we have shown that LTBQTE cannot overestimate the average
treatment effect in the lower tail.

The condition $\text{UTATE}(x) \leq \text{UTBQTE}(x)$ can be verified similarly.


### 1.3. Variance of the individual-level treatment effect

In this section, we consider how much the individual-level treatment effect
varies across the individuals and how that variance depends on the assumption
of order-preserving treatment. For that, we write the variance
of the individual-level treatment effect $X^* - X$ in terms of 
correlation $r = \text{cor}(X^*,X)$ and variance $v = \text{var}(X)$ of the untreated population
and $v^* = \text{var}(X^*)$ of the treated population as follows
\begin{eqnarray*}
\text{var}(X^*-X) &=& v + v^* - 2\,\text{cov}(X^*,X)\\
&=& v + v^* - 2\,r\, \sqrt{v\,v^*}.
\end{eqnarray*}
Thus, the variance of the treatment effect is the smallest possible when $r=1$,
in which case the treatment preserves order, and is the largest possible when 
$r=-1$, in which case the treatment completely flips the order around.

We conclude that the order-preserving treatment with
a linear relationship between the treated and untreated outcome values
is the most conservative example of treatment in the sense of
minimizing the variance of the treatment effect in the population.



# 2. Testing the `bqte( )` function to back-transform quantile treatment effect

Let's test `bqte()` function by generating data for the treatment ($T$) group
and the control ($C$) group.

**Continuous case.** For a given choice of continuous distributions, we can compute the exact 
BQTE for each possible outcome value in controls by first using
cumulative density function (CDF) in controls to get a correct tail probability 
in the control distribution
and then applying the quantile function (inverse of CDF) in the treatment group 
to get the corresponding outcome value in that group. 
Difference between the two outcome values is the exact BQTE.

**Discrete case.** For discrete distributions, the simulation setting is slightly more complex.
There, each possible outcome value in controls corresponds to a probability mass 
which in turn corresponds to a probability interval in CDF. 
When this interval is mapped by the inverse of CDF in the treatment group, 
the result contains possibly multiple outcome values in the treatment group.
To get the exact BQTE, we compute the expectation of the outcome value in the treatment group conditional 
on the event that corresponds to the particular interval of CDF. 
The difference between this expectation and the outcome value in the controls
is then the exact BQTE.

We apply three estimators of BQTE. The first two are versions of the estimator
presented in our main text and the third was defined by Doksum (1974).

 1. **Bagging estimator** that equals to the average of the BQTE estimates
 across the bootstrap samples. This is our default estimator and it is shown in Figures
 unless otherwise stated in the Figure caption.
 The motivation for bootstrap averaging ("bagging") is that it can reduce variance of the
 estimator in some contexts, see Fig. S4.

 2. **Direct estimator** where we apply our BQTE estimator to the original data 
 and not to the bootstrapped samples as in bagging. (In `bqte()`, this estimator
 can be chosen by setting `bagging = FALSE`.)
 
 3. **Doksum estimator** defined as $\widehat{\Delta}(x) = y_{i(x)} - x$ where index
 $i(x) = \left\lfloor n\,\widehat{F}(x)\right\rfloor + 1$ with $\widehat{F}$ being the empirical
 cumulative distribution function of the control group, $n$ the sample size of
 the treatment group and $y_i$ the value in the treatment group having rank $i$.

### 2.1 Generating continuous example data

To study the continuous case, let's assume 
$$T\sim \text{Weibull}(1.2,2) \text{ and } C \sim \text{Weibull}(3,5).$$ 
Figure S1 demonstrates this setting and shows the estimates of 
treatment effect from `bqte()` function.

```{r, echo = FALSE}
T.a = 1.2
T.b = 2
C.a = 3
C.b = 5
eval.Fig2 = FALSE # Simulations take time -- set FALSE to read results from a file
Fig2.filename = "bqte_simulations_1.2_2_3_5_N50_500.Rdata"
eval.Fig3 = FALSE # Simulations take time -- set FALSE to read results from a file
Fig3.filename = "bqte_simulations_1.2_2_3_5_N50_500_Kcoeff.Rdata" 
eval.Fig4 = FALSE # set FALSE to read results from a file
Fig4.filename = "bqte_simulations_Mossad_100iter.Rdata" 
```

```{r, fig.width = 9, fig,height = 8, echo = F}
x = seq(0.001, 10, 0.01)
par(mfrow = c(1,2))

#PANEL A
plot(x, dweibull(x, T.a, T.b), t = "l", col = "orange", lwd = 1.5, 
     xlab = "outcome", ylab = "density", ylim = c(0, 0.45), lty = 2)
grid()
lines(x, dweibull(x, C.a, C.b), col = "dodgerblue", lwd = 1.5)
legend("topright", col = c("orange", "dodgerblue"), 
       leg = c("treatment", "control"), lwd = 1.5, lty = c(2,1))
text(x = par("usr")[1] - 2, y = par("usr")[4] + 0.05,
     labels = c("A."), xpd = NA, cex = 2)

#PANEL B
x = qweibull(seq(0.001, 0.999, 0.001), C.a, C.b)
y = qweibull(seq(0.001, 0.999, 0.001), T.a, T.b) - x
plot(x, y, t = "l", lty = 1, lwd = 2, col = "limegreen", 
     xlab = "outcome in controls",
     ylab = "BQTE",
     xlim = c(0, 10), ylim = c(-3.5, 0))
grid()
set.seed(19)
N = 500
Trt = rweibull(N, T.a, T.b)
Con = rweibull(N, C.a, C.b)
res.bqte = bqte(Treatment = Trt, Control = Con, verbose = FALSE)
points(res.bqte$at, res.bqte$bqte, col = "navyblue", pch = 19)
arrows(res.bqte$at, res.bqte$bqte.low, res.bqte$at, res.bqte$bqte.up, 
       col = "navyblue", length = 0, code = 3)
legend("topright", col = c("limegreen","navyblue"), 
       lty = c(1,1), pch = c(NA,19),
       leg = c("exact","estimate (95%CI)"), lwd = 1.5)
text(x = par("usr")[1] - 2, y = par("usr")[4] + 0.4,
     labels = c("B."),xpd = NA, cex = 2)
```

**Figure S1.** Panel A shows the theoretical distributions in the two groups:
$\text{Weibull}(1.2,2)$ in the treatment group and $\text{Weibull}(3,5)$ in the control group.
Panel B shows the exact back-transformed quantile treatment effect (BQTE)
as a function of the outcome value in controls (green line), 
together with estimates and their 95% CIs when 
`bqte()` was applied on a simulated data set 
with $N = 500$ individuals in each group.
\vspace{0.5cm}


### 2.2. The range of valid estimates

By default, `bqte()` estimates BQTE in the interval 
between the empirical quantile levels of 
$\frac{5}{N}$ and $\frac{N-5}{N}$, where $N$ is
the sample size in the control group. 
For example, in Fig. S1B, $N=500$ and hence
the range of estimation is between the 1st and the 99th percentile 
of the observed data, which correspond to the values of
2 and 8, respectively, when rounded to integers.

Let's confirm that the default range for estimation leads to well
calibrated confidence intervals (CIs), and let's also check that the estimates at 
more extreme quantile levels of $\frac{1}{N}$ and $\frac{N-1}{N}$ 
would not anymore remain correctly calibrated.

For both values of $N = 50$ and $N = 500$, we generate $R = 100$ data sets 
as above from the two Weibull distributions and apply `bqte()` function
to estimate BQTE 
at the outcome values that correspond to the following 5 quantile levels
in controls:
$\frac{1}{N}, \frac{5}{N}, \frac{1}{2}, \frac{N-5}{N}, \frac{N-1}{N}.$
For each $N$ and each quantile level, we then compute 
the empirical coverage of the 95%CIs
and the empirical mean of the BQTE estimates 
over the $R$ data sets. Results are show in Fig S2.

```{r, echo = FALSE}
qte.conf = 0.95
Ns = c(50, 500)
R = 100
```

```{r, eval = eval.Fig2, echo = FALSE}
set.seed(121)
res = list()
for(ii in 1:length(Ns)){
  N = Ns[ii]
  p.quant = c(1/N, 5/N, 0.5, 1-5/N, 1-1/N)
  lab.quant = c("1/N", "5/N", "1/2", "1-5/N", "1-1/N")
  at = qweibull(p.quant, C.a, C.b)
  #exact values for treatment effects 'at' points
  exact.bqte = qweibull(pweibull(at, C.a, C.b), T.a, T.b) - at
  inside = rep(0, length(at))
  est = matrix(NA, ncol = length(at), nrow = R)
  est.direct = matrix(NA, ncol = length(at), nrow = R)
  est.doksum = matrix(NA, ncol = length(at), nrow = R)
  
  for(set in 1:R){
    Trt = rweibull(N, T.a, T.b)
    Con = rweibull(N, C.a, C.b)
    
    # bagging estimate with confidence intervals
    res.bqte = bqte(Treatment = Trt, Control = Con, 
                    at = at, verbose = FALSE, B = 2000,
                    bagging = TRUE)
    inside = inside + as.numeric((res.bqte$bqte.low < exact.bqte & 
                                    res.bqte$bqte.up > exact.bqte))
    est[set,] = res.bqte$bqte
    
    # direct estimate without bagging (and no confidence intervals)
    res.bqte = bqte(Treatment = Trt, Control = Con, 
                    at = at, verbose = FALSE, B = 1,
                    bagging = FALSE)
    est.direct[set,] = res.bqte$bqte
    
    # Doksum estimate
    est.doksum[set,] = bqte.doksum(at = at, Treatment = Trt, Control = Con)
  }
  mu = apply(est, 2, mean)
  se = apply(est, 2, sd) / sqrt(R)
  rmse = sqrt(rowSums((t(est) - exact.bqte)^2)/R)
  mu.direct = apply(est.direct, 2, mean)
  se.direct = apply(est.direct, 2, sd) / sqrt(R)
  rmse.direct = sqrt(rowSums((t(est.direct) - exact.bqte)^2)/R)
  mu.doksum = apply(est.doksum, 2, mean)
  se.doksum = apply(est.doksum, 2, sd) / sqrt(R)
  rmse.doksum = sqrt(rowSums((t(est.doksum) - exact.bqte)^2)/R)

  ci = sapply(inside, function(x){ binom.test(x, R, conf.level = 0.95)$conf.int})
  rel.bias = (mu - exact.bqte) / abs(exact.bqte) 
  res[[ii]] = data.frame(label = lab.quant, at, 
                         exact = exact.bqte, 
                         est = mu, 
                         se.est = se, 
                         rmse = rmse,
                         rel.bias, 
                         est.direct = mu.direct, 
                         se.est.direct = se.direct, 
                         rmse.direct = rmse.direct,
                         est.doksum = mu.doksum, 
                         se.est.doksum = se.doksum, 
                         rmse.doksum = rmse.doksum,
                         ci.cov = inside/R, 
                         ci.cov.low = ci[1,], 
                         ci.cov.up = ci[2,],
                         est.direct = apply(est, 2, mean)
                        )
}
params = data.frame(T.a, T.b, C.a, C.b)
save(res, Ns, params, file = Fig2.filename)
```


```{r, fig.width = 9, fig.height = 3, echo = FALSE}
if(!eval.Fig2) load(Fig2.filename)
par(mfrow = c(1, 3))
par(mar = c(5.5, 5.5, 2, 1))
n.quant = nrow(res[[1]])
plot(NULL, xlim = c(0.5, n.quant + 0.5), ylim = c(60, 100), 
     xaxt = "n", yaxt = "n", xlab = "quantile level", 
     ylab = "coverage (%) of 95%CI", cex.lab = 1.3)
adj = -0.1
cols = c("orange","dodgerblue")
pchs = c(0, 1, 5)
grid()
abline(h = 95)

for(ii in 1:length(Ns)){
  points(1:n.quant + adj, res[[ii]]$ci.cov*100,
         col = cols[ii], pch = pchs[ii], cex = 1.2, lwd = 1.6)
  arrows(1:n.quant + adj, res[[ii]]$ci.cov.low*100,
         1:n.quant + adj, res[[ii]]$ci.cov.up*100, 
         col = cols[ii], length = 0, code = 3, lwd = 1.2)
  adj = adj + 0.2
}
#Add legend manually
rect(xleft = 4.0, ybottom = 58, xright = 5.7, ytop = 70,
     col = "white", border = NULL)
points(rep(4.3, 2), c(61, 67), pch = pchs, 
       col = cols, lwd = 1.5, cex = 1.2)
text(c(5, 5.1), c(61, 67), labels = paste0("N=",Ns), cex  = 1.3)

axis(1, at = 1:n.quant, labels = FALSE)
text(x = 1:n.quant, y = par("usr")[3] - 3.8, res[[1]]$label,
     xpd = NA, srt = 35, adj = 0.7, cex = 1.3)
axis(2, at = c(60, 70, 80, 90, 95, 100), cex.axis = 1.3, las = 1)
text(x = par("usr")[1] - 1.2, y = par("usr")[4] + 3,
     labels = c("A."),xpd = NA, cex = 2)

plot(NULL, xlim = c(0.5, n.quant + 0.5), ylim = c(-1,1), 
     xaxt = "n", yaxt = "n", xlab = "quantile level", 
     ylab = "bias", cex.lab = 1.3)
axis(1, at = 1:n.quant, labels = FALSE)
text(x = 1:n.quant, y = par("usr")[3] - 0.2,
     res[[ii]]$label,
     xpd = NA, srt = 35, adj = 0.7, cex = 1.3)
axis(2, at = c(-1, -0.5, 0, 0.5, 1), cex.axis = 1.3, las = 1)
adj = -0.1
grid()
abline(h = 0)

for(ii in 1:length(Ns)){
  bias = res[[ii]]$est - res[[ii]]$exact
  se = res[[ii]]$se.est 
  points(1:n.quant + adj, bias,
         col = cols[ii], pch = pchs[ii], cex = 1.2, lwd = 1.6)
  arrows(1:n.quant + adj, bias + 1.96*se,
         1:n.quant + adj, bias - 1.96*se, 
         col = cols[ii], length = 0, 
         code = 3, lwd = 1.2)
  adj = adj + 0.2
}
text(x = par("usr")[1] - 1.2, y = par("usr")[4] + 0.15,
     labels = c("B."), xpd = NA, cex = 2)

x = qweibull(seq(0.001, 0.999, 0.001), C.a, C.b)
y = qweibull(seq(0.001, 0.999, 0.001), T.a, T.b) - x
plot(x, y, t = "l", lty = 1, lwd = 2, col = "limegreen", 
     xlab = "outcome in controls", xlim = c(0, 10), ylim = c(-3.5, 0),
     ylab = "BQTE", cex.lab = 1.3, cex.axis = 1.3, las = 1)
grid()

for(ii in 1:length(Ns)){
  bias = res[[ii]]$est - res[[ii]]$exact
  se = res[[ii]]$se.est 
  points(res[[ii]]$at, res[[ii]]$est,
         col = cols[ii], pch = pchs[ii], cex = 1.2, lwd = 1.6)
  arrows(res[[ii]]$at, res[[ii]]$est + 1.96*se,
         res[[ii]]$at, res[[ii]]$est - 1.96*se, 
         col = cols[ii], length = 0, 
         code = 3, lwd = 1.2)
}
text(x = par("usr")[1] - 2.5, y = par("usr")[4] + 0.25,
     labels = c("C."), xpd = NA, cex = 2)
```

**Figure S2.** Observed coverage of 95% confidence intervals (panel A),
observed bias (panel B) and
the mean of the estimates over the $100$ replicates compared to the 
exact values shown by the green line (panel C). 
The value of $N$ is shown in the legend of panel A.
The intervals around each point show 95% confidence intervals 
based on the sample size of $100$ replicates.
\vspace{0.5cm}


In Fig. S2A we see that, independent of sample size $N$, 
the 95%CIs are well calibrated for quantile levels $\frac{5}{N}$ and $\frac{N-5}{N}$ 
but not for more extreme levels of $\frac{1}{N}$ or $\frac{N-1}{N}$. 
Fig. S2B shows that there is more bias at the extreme quantile levels 
than near the median for both sample sizes and the bias does not depend much
on the sample size except at the extreme quantile levels. 
(Note, however, that for any fixed outcome value in controls, 
the bias would disappear as sample size grows;
this is demonstrated by the estimates at the median point in panel B
where the bias disappears as $N$ grows from 50 to 500.)
From Fig. S2C we observe how these biases correspond to some overestimation
of the magnitude of the quantile treatment effect (QTE) at the extreme quantile levels of 
$\frac{1}{N}$ and $\frac{N-1}{N}$.
Based on these results, 
our default option is to restrict the estimation of BQTE and QTE
between quantile levels of $\frac{5}{N}$ and $\frac{N-5}{N}$.

Figure S2 shows only the bagging estimator but we also evaluated the other two
estimators on the same simulation setting. 
The root mean square error (RMSE) of the three estimators at the quantile levels of 
$5/N$, $1/2$ and $1-5/N$ were, for the sample size of 50:
```{r, echo = FALSE}
#Compare 3 estimators in terms of RMSE
data.frame(quantile = c("5/N","1/2","1-5/N"), 
           at = res[[1]]$at[2:4], 
           RMSE_bagging = res[[1]]$rmse[2:4],
           RMSE_direct = res[[1]]$rmse.direct[2:4],
           RMSE_doksum = res[[1]]$rmse.doksum[2:4])
```
and for the sample size of 500:
```{r, echo = FALSE}
data.frame(quantile = c("5/N","1/2","1-5/N"),
           at = res[[2]]$at[2:4], 
           RMSE_bagging = res[[2]]$rmse[2:4],
           RMSE_direct = res[[2]]$rmse.direct[2:4],
           RMSE_doksum = res[[2]]$rmse.doksum[2:4])
```


### 2.3. The number of quantile points used in estimation

The `bqte()` function takes as input the 
number $K$ of quantile points, corresponding to the probability values of
$1/(K+1),2/(K+1),\ldots,K/(K+1)$, that are used in the estimation of
BQTEs. Between these points, BQTE is estimated
by linear interpolation.
By default, we set $K = N$, where $N$ is the size of the control group.
The value $N$ is a natural upper bound for $K$ because when $K = N$,
there is already a separate estimate corresponding to each control observation.
But could some smaller value for $K < N$ be a better choice?
Let's use a similar simulation procedure as in Fig. S2 to assess how
smaller values of $K < N$ compare to our default choice $K=N.$

For this simulation, the sample size in both groups 
takes values $N = 50$ or $N = 500$, and for both values of $N$,
we will run 100 simulations for each value of 
$K = \frac{N}{4}, \frac{N}{2}, \frac{2N}{3}, N$.
The results are in Fig. S3.

```{r, echo = FALSE}
Ns = c(50, 500)
Kcoeffs = c(0.25, 0.5, 0.66, 1)
R = 100
```

```{r, eval = eval.Fig3, echo = FALSE}
set.seed(123)
res = list()
res.ii = 1
for(kk in 1:length(Ns)){
  for(ii in 1:length(Kcoeffs)){
    N = Ns[kk]
    Kc = Kcoeffs[ii]
    p.quant = c(1/N, 5/N, 0.5, 1-5/N, 1-1/N)
    lab.quant = c("1/N", "5/N", "1/2", "1-5/N", "1-1/N")
    at = qweibull(p.quant, C.a, C.b)
  #exact values for BQTEs at points 'at'
    exact.bqte = qweibull(pweibull(at, C.a, C.b), T.a, T.b) - at
    inside = rep(0, length(at)) #coverage
    est = matrix(NA, ncol = length(at), nrow = R)
    for(set in 1:R){
      Trt = rweibull(N, T.a, T.b)
      Con = rweibull(N, C.a, C.b)
      res.bqte = bqte(Treatment = Trt, Control = Con, 
                      at = at, verbose = FALSE, K = ceiling(N*Kc))
      inside = inside + as.numeric((res.bqte$bqte.low < exact.bqte & 
                                    res.bqte$bqte.up > exact.bqte))
      est[set,] = res.bqte$bqte
    }
    mu = apply(est, 2, mean)
    se = apply(est, 2, sd)/sqrt(R)
    ci = sapply(inside, function(x){ binom.test(x, R, conf.level = 0.95)$conf.int})
    rel.bias = (mu - exact.bqte)/abs(exact.bqte) 
    res[[res.ii]] = data.frame(labels = lab.quant, at, exact = exact.bqte,
                               est = mu, se.est = se, rel.bias, 
                               ci.cov = inside/R, 
                               ci.cov.low = ci[1,], 
                               ci.cov.up = ci[2,])
    res.ii = res.ii + 1
  }
}
params = data.frame(T.a, T.b, C.a, C.b)
save(res, Ns, Kcoeffs, params, file = Fig3.filename)
```


```{r, fig.width = 9.5, fig.height = 6, echo = FALSE}
if(!eval.Fig3) load(Fig3.filename)
par(mfrow = c(length(Ns), 3))
par(mar = c(5.5, 5.5, 2, 1))
n.quant = nrow(res[[1]])
res.ii = 0
x = qweibull(seq(0.001, 0.999, 0.001), C.a, C.b)
y = qweibull(seq(0.001, 0.999, 0.001), T.a, T.b) - x

for(N.ii in 1:length(Ns)){
  N.C = Ns[N.ii]
  plot(NULL, xlim = c(0.5, n.quant + 0.5), ylim = c(40, 100), 
       xaxt = "n", yaxt = "n", xlab = "quantile level", 
       ylab = "coverage (%) of 95%CI", cex.lab = 1.3, 
       main = substitute(N==N.C, list(N.C = N.C)), cex.main = 1.3)
  adj = -0.3
  cols = c("red", "orange", "dodgerblue", "navyblue")
  pchs = c(0, 1, 5, 6)
  grid()
  abline(h = 95)

  for(ii in 1:length(Kcoeffs)){
    points(1:n.quant + adj, res[[res.ii + ii]]$ci.cov*100,
           col = cols[ii], pch = pchs[ii], cex = 1.2, lwd = 1.6)
    arrows(1:n.quant + adj, res[[res.ii +ii]]$ci.cov.low*100,
           1:n.quant + adj, res[[res.ii +ii]]$ci.cov.up*100, 
           col = cols[ii], length = 0, code = 3, lwd = 1.2)
    adj = adj + 0.2
  }
  #Add legend manually
  rect(xleft = 3.5, ybottom = 37, xright = 5.7, ytop = 60,
       col = "white", border = NULL)
  points(rep(3.7, 4),c(40, 46, 52, 58), 
         pch = pchs, col = cols, lwd = 1.5, cex = 1.2)
  text(c(4.6, 4.5, 4.6, 4.4), c(40, 46, 52, 58), 
       labels = paste0("K/N=",Kcoeffs), cex  = 1.3)

  axis(1, at = 1:n.quant, labels = FALSE)
  text(x = 1:n.quant, y = par("usr")[3] - 5.5,
       res[[1]]$labels,
       xpd = NA, srt = 35, adj = 0.7, cex = 1.3)
  axis(2, at = c(40, 50, 60, 70, 80, 90, 100), cex.axis = 1.3, las = 2)
  text(x = par("usr")[1] - 1.2, y = par("usr")[4] + 5,
       labels = c("A.","D.","G.")[N.ii],xpd = NA, cex = 2)

  plot(NULL, xlim = c(0.5,n.quant + 0.5), ylim = c(-1,1), 
       xaxt = "n", yaxt = "n", xlab = "quantile level", 
       ylab = "bias", cex.lab = 1.3, 
       main = substitute(N==N.C, list(N.C = N.C)), cex.main = 1.3)
  axis(1, at = 1:n.quant, labels = FALSE)
  text(x = 1:n.quant, y = par("usr")[3] - 0.2,
       res[[1]]$labels,
       xpd = NA, srt = 35, adj = 0.7, cex = 1.3)
  axis(2, at = c(-1, -0.5, 0, 0.5, 1), cex.axis = 1.3, las = 2)
  adj = -0.3
  grid()
  abline(h = 0)

  for(ii in 1:length(Kcoeffs)){
    bias = res[[res.ii + ii]]$est - res[[res.ii + ii]]$exact
    se = res[[res.ii + ii]]$se.est 
    points(1:n.quant + adj, bias,
         col = cols[ii], pch = pchs[ii], cex = 1.2, lwd = 1.6)
    arrows(1:n.quant + adj, bias + 1.96*se,
         1:n.quant + adj, bias - 1.96*se, 
         col = cols[ii], length = 0, 
         code = 3, lwd = 1.2)
    adj = adj + 0.2
  }
  text(x = par("usr")[1] - 1.2, y = par("usr")[4] + 0.15,
       labels = c("B.","E.","H.")[N.ii], xpd = NA, cex = 2)

  plot(x, y, t = "l", lty = 1, lwd = 2, col = "limegreen", 
       xlab = "outcome in controls", xlim = c(0,10), ylim = c(-3.5,0),
       ylab = "BQTE", cex.lab = 1.3, cex.axis = 1.3, las = 1,
       main = substitute(N==N.C, list(N.C = N.C)), cex.main = 1.3)
  grid()
  for(ii in 1:length(Kcoeffs)){
    bias = res[[res.ii + ii]]$est - res[[res.ii + ii]]$exact
    se = res[[res.ii + ii]]$se.est 
    points(res[[res.ii +ii]]$at, res[[res.ii + ii]]$est,
           col = cols[ii], pch = pchs[ii], cex = 1.2, lwd = 1.6)
    arrows(res[[res.ii + ii]]$at, res[[res.ii + ii]]$est + 1.96*se,
           res[[res.ii + ii]]$at, res[[res.ii + ii]]$est - 1.96*se, 
           col = cols[ii], length = 0, 
           code = 3, lwd = 1.2)
  }
  text(x = par("usr")[1] - 2.5, y = par("usr")[4] + 0.25,
       labels = c("C.","F.","I.")[N.ii],xpd = NA, cex = 2)
  res.ii = res.ii + length(Kcoeffs)
}
```

**Figure S3.** Observed coverage of 95% confidence intervals (panels A, D),
observed bias (panels B, E) and
the mean of the estimates over the $100$ replicates compared to the 
exact values shown by the green line (panels C, F). 
The ratio $K/N$ is shown in the legend of panels A and D, 
and $N$ is shown in the title of each panel. The intervals around each 
point show 95% confidence intervals based on the sample size of $100$ replicates.
\vspace{0.5cm}

In Figure S3, we do not observe clear difference between different values of $K/N$
except at the extreme quantile levels of $1/N$ and $1-1/N$,
and hence the choice $K = N$ seems a suitable default 
value to be used in `bqte()` function.


### 2.4. Magnitude of estimation error and coverage of confidence intervals in discrete data 

We used the Mossad data from the main paper to generate realistic data about the
duration of cold in a randomized trial when the duration is measured in days. 
We focus on the recommended interval between quantile levels $5/N$ and $1-5/N$ that
in these data corresponds to interval (3,17).
For each integer value in this interval, we computed the exact BQTE values for the 
observed distribution of the Mossad data using function `exact.bqte.discrete()`.
Then we generated 100 simulations from the empirical distribution from Mossad data 
with sample sizes equaling to the Mossad data (Treatment group = 49, Control group = 50).
We estimated BQTE by all three estimators.
Figure S4 shows the root mean squared error of the three estimators
as well as the coverage of the 95% confidence interval by our
bootstrap method.


```{r, eval = eval.Fig4, echo = FALSE}
# Data set 1, Mossad data
filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/Mossad.csv"
dat <- read.csv(filename)
duration = dat$Duration
status = dat$Zinc

x = seq(1, max(duration))
#Empirical probability distributions for two groups:
#Treatment group
tab = table(duration[status == 1])
pr.t = rep(0, length(x))
pr.t[as.numeric(names(tab))] = as.numeric(tab)
pr.t = pr.t / sum(pr.t)
#Control group
tab = table(duration[status == 0])
pr.c = rep(0, length(x))
pr.c[as.numeric(names(tab))] = as.numeric(tab)
pr.c = pr.c / sum(pr.c)

# Compute recommended range for BQTE estimation
n.c = sum(status == 0)
n.t = sum(status == 1)
at = c(quantiles.from.density(p = 5/n.c, pr.c), 
       quantiles.from.density(p = 1 - 5/n.c, pr.c))
at = seq(at[1], at[2])
# Compute true BQTE for current data
true.bqte = sapply(at, function(a){exact.bqte.discrete(a, x, pr.c, pr.t)})

# Compare performance of three estimators of BQTE:
# 1. Direct estimate from data (without bagging)
# 2. Bagging estimate (average of the bootstrap distribution)
# 3. Doksum estimate 
n.iter = 100
res.dir = matrix(NA, ncol = length(at), nrow = n.iter)
res.bag = matrix(NA, ncol = length(at), nrow = n.iter)
res.doksum = matrix(NA, ncol = length(at), nrow = n.iter)
within.ci = rep(0, length(at)) # To estimate the coverage of CIs from bootstrap
set.seed(10)
for(ii in 1:n.iter){
  #Generate data from pr.c (Controls) and pr.t (Treated)
  Control = sample(x, size = n.c, replace = T, prob = pr.c)
  Treatment = sample(x, size = n.t, replace = T, prob = pr.t)
  
  #Without bagging (Ignores bootstrap and sets B=1 to speed up)
  res = bqte(Treatment, Control, at = at, 
             bqte.conf = 0.95, B = 1, 
             K = length(Control), bagging = FALSE,
             verbose = FALSE)
  res.dir[ii,] = res$bqte
  
  #With bagging
  res = bqte(Treatment, Control, at = at, 
             bqte.conf = 0.95, B = 2000, 
             K = length(Control), bagging = TRUE,
             verbose = FALSE)
  res.bag[ii,] = res$bqte
  
  within.ci = within.ci + as.numeric(res$bqte.low <= true.bqte & res$bqte.up >= true.bqte)
  
  #Doksum estimator
  res = bqte.doksum(at = at, Treatment, Control)
  res.doksum[ii,] = res
}

rmse.dir = sqrt(apply( (t(res.dir) - true.bqte)^2, 1, mean))
rmse.bag = sqrt(apply( (t(res.bag) - true.bqte)^2, 1, mean))
rmse.doksum = sqrt(apply( (t(res.doksum) - true.bqte)^2, 1, mean))
params = data.frame(n.iter = n.iter)
res = data.frame(at = at, 
                 true.bqte = true.bqte, 
                 rmse.dir = rmse.dir, 
                 rmse.bag = rmse.bag, 
                 rmse.doksum = rmse.doksum, 
                 within.ci = within.ci/n.iter)
save(res, params, file = Fig4.filename)
```

```{r, fig.width = 10, fig.height = 5, echo = FALSE}
if(!eval.Fig4) load(Fig4.filename)
par(mfrow = c(1,2))
plot(res$at, res$rmse.dir, t = "l", col = "red", lty = 2, 
     ylim = c(0, max(c(res$rmse.dir, res$rmse.bag, res$rmse.doksum))), 
     xlab = "outcome in controls", ylab = "root mean squared error")
lines(res$at, res$rmse.bag, col = "blue", t = "l", lty = 1)
lines(res$at, res$rmse.doksum, col = "orange", t = "l", lty = 4)

legend("bottomright", leg = c("direct","bagging","Doksum"), 
       col = c("red","blue","orange"), lty = c(2,1,4))

text(x = par("usr")[1] - 2, y = par("usr")[4] + 0.05,
     labels = c("A."), xpd = NA, cex = 2)

ci = t(sapply(round(params$n.iter*res$within.ci), function(y){as.numeric(
  binom.test(y, n = params$n.iter, conf.level = 0.95)$conf.int)} ))
plot(res$at, 100*res$within.ci, xlab = "outcome in controls", ylab = "coverage (%) of 95%CI", 
     ylim = c(85, 100), col = "white")
abline(h = 95, lty = 2, col = "red")
arrows(res$at, 100*ci[,1], res$at, 100*ci[,2], col = "gray", lwd = 2, code = 3, length = 0, angle = 90)
points(res$at, 100*res$within.ci, col = "black", pch = 19)
text(x = par("usr")[1] - 2.2, y = par("usr")[4] + 0.4,
     labels = c("B."),xpd = NA, cex = 2)
```

**Figure S4.** Panel A shows root mean squared error (RMSE) for three estimators 
(bagging in blue, direct in red and Doksum in orange) at each of the integer valued cold duration
between 3 and 17 days. The RMSE is estimated over 100 simulated data sets that follow the
observed distributions of the Doksum data from the main text.
Panel B shows the empirical coverage of the 95% confidence intervals estimated with bootstrap
at the same grid of values as in panel A. The intervals around the point estimates are 95%
binomial confidence intervals based on 100 samples and the red line denotes the target coverage
of 95%. 
\vspace{0.5cm}

From Figure S4 we see that the RMSE seems the smallest for the bagging estimator. 
As in the continuous setting, there were little differences between the three estimators, 
we assume that the differences seen here are due to the discreteness of the data. 
Based on the result of Figure S4A, we have used the
bagging estimator as our default estimator in the discrete data sets that we have analyzed in the main text.

Panel B shows that the 95% confidence intervals do not show any evidence of falling short of the 
target coverage of 95% as the confidence intervals of the coverage estimates 
in panel B either overlap with the target value or remain above the target value in every case.


\newpage
# 3. Source of data of the included examples

**Mossad study.** The data are from pages 2-4 of the Additional file 2 of

<https://doi.org/10.1186/s12874-017-0356-y>.

**Zinc acetate studies.** The data are from page 8 of the Supplementary file 2 of

<https://doi.org/10.1093/ofid/ofx059>

**Carrageenan studies.** The data are from page 3 of the Supplementary file of

<https://doi.org/10.1002/prp2.810>
\newpage


# 4. Supplementary figures of real data sets

```{r echo = F}
for(data.set in c(1)){
  # Data set 1, Mossad data
  if(data.set == 1){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/Mossad.csv"
    filename = "Mossad.csv"
    title.lab = "Mossad"
    yrange.d = c(-10, 0)
    yrange.r = c(-80,0)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Zinc
  }
  
  # Data set 2, Zinc acetat data
  if(data.set == 2){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/ZnAcet.csv"
    filename = "ZnAcet.csv"
    title.lab = "Zinc acetat"
    yrange.d = c(-8, 0.5)
    yrange.r = c(-60,0)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Zinc
  }
  
  # Data set 3, carrageenan data
  if(data.set == 2){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/Carrageenan.csv"
    filename = "Carrageenan.csv"
    title.lab = "Carrageenan"
    yrange.d = c(-10, 1.5)
    yrange.r = c(-70,30)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Carr
  }

Treatment = duration[status == 1]
Control = duration[status == 0]

res = bqte(Treatment, Control, at = NULL, 
           bqte.conf = 0.95, B = 2000, 
           K = length(Control), bagging = TRUE,
           tails = TRUE,
           verbose = FALSE)
#res

xlab = "Days in controls"
pch.sty = NULL
par(mfrow = c(2,3))
for(rel.scale in c(FALSE,TRUE)){

  ate = mean(Treatment) - mean(Control)
  txt = title.lab
  yrange = yrange.d
  
  if(rel.scale) {
    ate = 100*ate/mean(Control)
    txt = ""
    yrange = yrange.r
  }
  plot_bqte(res, plot.ci = TRUE, plot.rbqte = rel.scale,
            pch = pch.sty, col = "black", col.ci = "gray",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            main = txt, cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")

  plot_tbqte(res, utbqte = FALSE, plot.ci = TRUE, plot.relative = rel.scale,
            pch = pch.sty, col.ci = "lightblue", col = "black",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")

  plot_tbqte(res, utbqte = TRUE, plot.ci = TRUE, plot.relative = rel.scale,
            pch = pch.sty, col.ci = "salmon", col = "black",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")
  }
}
```

**Figure S5.** BQTE, LTBQTE and UTBQTE functions for the Mossad data set. 
The top row shows the values on the outcome scale.
The bottom row shows the values on the relative scale. 
The solid black lines shows the estimates and the band shows the 95% 
confidence interval based on the bootstrap samples.
The horizontal dashed red lines mark the average treatment effect.
\vspace{0.5cm}

```{r echo = F}
for(data.set in c(2)){
  # Data set 1, Mossad data
  if(data.set == 1){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/Mossad.csv"
    filename = "Mossad.csv"
    title.lab = "Mossad"
    yrange.d = c(-10, 0)
    yrange.r = c(-80,0)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Zinc
  }
  
  # Data set 2, Zinc acetat data
  if(data.set == 2){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/ZnAcet.csv"
    filename = "ZnAcet.csv"
    title.lab = "Zinc acetat"
    yrange.d = c(-8, 0.5)
    yrange.r = c(-60,0)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Zinc
  }
  
  # Data set 3, carrageenan data
  if(data.set == 3){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/Carrageenan.csv"
    filename = "Carrageenan.csv"
    title.lab = "Carrageenan"
    yrange.d = c(-10, 1.5)
    yrange.r = c(-70,30)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Carr
  }

Treatment = duration[status == 1]
Control = duration[status == 0]

res = bqte(Treatment, Control, at = NULL, 
           bqte.conf = 0.95, B = 2000, 
           K = length(Control), bagging = TRUE,
           tails = TRUE,
           verbose = FALSE)
#res

xlab = "Days in controls"
pch.sty = NULL
par(mfrow = c(2,3))
for(rel.scale in c(FALSE,TRUE)){

  ate = mean(Treatment) - mean(Control)
  txt = title.lab
  yrange = yrange.d
  
  if(rel.scale) {
    ate = 100*ate/mean(Control)
    txt = ""
    yrange = yrange.r
  }
  plot_bqte(res, plot.ci = TRUE, plot.rbqte = rel.scale,
            pch = pch.sty, col = "black", col.ci = "gray",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            main = txt, cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")

  plot_tbqte(res, utbqte = FALSE, plot.ci = TRUE, plot.relative = rel.scale,
            pch = pch.sty, col.ci = "lightblue", col = "black",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")

  plot_tbqte(res, utbqte = TRUE, plot.ci = TRUE, plot.relative = rel.scale,
            pch = pch.sty, col.ci = "salmon", col = "black",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")
  }
}
```

**Figure S6.** BQTE, LTBQTE and UTBQTE functions for the Zinc acetat data set. 
The top row shows the values on the outcome scale.
The bottom row shows the values on the relative scale. 
The solid black lines shows the estimates and the band shows the 95% 
confidence interval based on the bootstrap samples.
The horizontal dashed red lines mark the average treatment effect.
\vspace{0.5cm}


```{r echo = F}
for(data.set in c(3)){
  # Data set 1, Mossad data
  if(data.set == 1){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/Mossad.csv"
    filename = "Mossad.csv"
    title.lab = "Mossad"
    yrange.d = c(-10, 0)
    yrange.r = c(-80,0)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Zinc
  }
  
  # Data set 2, Zinc acetat data
  if(data.set == 2){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/ZnAcet.csv"
    filename = "ZnAcet.csv"
    title.lab = "Zinc acetat"
    yrange.d = c(-8, 0.5)
    yrange.r = c(-60,0)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Zinc
  }
  
  # Data set 3, carrageenan data
  if(data.set == 3){
    #filename = "https://raw.githubusercontent.com/mjpirinen/qte-by-outcome/main/Carrageenan.csv"
    filename = "Carrageenan.csv"
    title.lab = "Carrageenan"
    yrange.d = c(-10, 1.5)
    yrange.r = c(-70,30)
    dat = read.csv(filename)
    duration = dat$Duration
    status = dat$Carr
  }

Treatment = duration[status == 1]
Control = duration[status == 0]

res = bqte(Treatment, Control, at = NULL, 
           bqte.conf = 0.95, B = 2000, 
           K = length(Control), bagging = TRUE,
           tails = TRUE,
           verbose = FALSE)
#res

xlab = "Days in controls"
pch.sty = NULL
par(mfrow = c(2,3))
for(rel.scale in c(FALSE,TRUE)){

  ate = mean(Treatment) - mean(Control)
  txt = title.lab
  yrange = yrange.d
  
  if(rel.scale) {
    ate = 100*ate/mean(Control)
    txt = ""
    yrange = yrange.r
  }
  plot_bqte(res, plot.ci = TRUE, plot.rbqte = rel.scale,
            pch = pch.sty, col = "black", col.ci = "gray",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            main = txt, cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")

  plot_tbqte(res, utbqte = FALSE, plot.ci = TRUE, plot.relative = rel.scale,
            pch = pch.sty, col.ci = "lightblue", col = "black",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")

  plot_tbqte(res, utbqte = TRUE, plot.ci = TRUE, plot.relative = rel.scale,
            pch = pch.sty, col.ci = "salmon", col = "black",
            xlim = NULL, ylim = yrange, 
            xlab = xlab, ylab = NULL,
            xaxs = "r", yaxs = "r",
            xaxp = NULL, yaxp = NULL,
            cex.axis = 1.3, cex.lab = 1.2)
  abline(h = ate, lty = 2, lwd = 1, col = "red")
  }
}
```

**Figure S7.** BQTE, LTBQTE and UTBQTE functions for the Carrageenan data set. 
The top row shows the values on the outcome scale.
The bottom row shows the values on the relative scale. 
The solid black lines shows the estimates and the band shows the 95% 
confidence interval based on the bootstrap samples.
The horizontal dashed red lines mark the average treatment effect.
\vspace{0.5cm}

```{r, echo = F, fig.width = 6, fig.height = 6}
# Comparison of the Carrageenan data with different 
# imputation of the censored values imputation 20 days 
# Only difference between the data sets is that 
# the censored observations on day 20 have been imputed to either 20 or to 30 days.

filename = "https://raw.githubusercontent.com/mjpirinen/bqte/main/Carrageenan.csv"
Carrageenan20 = read.csv(filename) #here censored values have been imputed as 20
Treatment20 = subset(Carrageenan20, Carr == 1)$Duration
Control20 = subset(Carrageenan20, Carr == 0)$Duration

filename = "https://raw.githubusercontent.com/mjpirinen/bqte/main/Carrageenan30.csv"
Carrageenan30 = read.csv(filename) #here censored values have been imputed as 30
Treatment30 = subset(Carrageenan30, Carr == 1)$Duration
Control30 = subset(Carrageenan30, Carr == 0)$Duration

set.seed(4)
#Use more bootstrap samples to ensure max accuracy (20000 while default = 2000)
res20.bqte = bqte(Treatment = Treatment20, Control = Control20, 
                B = 20000, at = c(3:19), verbose = F) 
res30.bqte = bqte(Treatment = Treatment30, Control = Control30, 
                  B = 20000, at = c(3:19), verbose = F)

par(mfrow = c(1,1))
plot (res20.bqte$at, res20.bqte$bqte, type ="l", xlim = c(0,20), 
      yaxp = c(-10, 0, 5), xaxs = "i",
      ylab = "BQTE (Days)", 
      xlab = "Days in controls")
lines(res30.bqte$at, res30.bqte$bqte, lty = 2, col = "red")
legend("bottomleft", col = c("black","red"), lty = c(1,2), 
       leg = c("cens. to 20", "cens. to 30"))
# Conclusion: 
# No difference observed due to imputation in the range considered here.
```


**Figure S8.** Effect of imputation of the censored observations in the Carrageenan data set. The continuous black line shows the BQTE function when the censored observations
were imputed using the value of 20 days. 
The dashed red line shows the BQTE function when the censored observations
were imputed using the value of 30 days. In this calculation, we used B = 20 000 
bootstrap samples to ensure stable estimates. 
There is only a very small difference between 
the curves at the right-hand side, 
indicating that the curves are robust to the type of imputation applied in this case.  
